---
title: "00_Leukemia_in_NY"
author: "Frances Lin"
date: "3/23/2022"
output: pdf_document
---

This project fits 6 models (fixed effects, random effects (iid), SLM, ICAR, BYM and Leroux et al.), produces summary results and plots results. 

## Load packages

```{r}
# Load packages
library(spdep)     # for spatial weights matrix objects
library(DClusterm) # for data
library(tidyverse)
library(pander)
library(ggplot2)
library(gridExtra)
library(sjmisc)    # transpose df 
library(here)
```

```{r}
# Load data
library(DClusterm)
data(NY8)
```


## The `NY8` data

The `NY8` data set contains the number of incident leukemia cases per census tract in an eight-country region of upstate New York from 1978-1982 (Waller & Gotway, 2004; Bivand et al., 2008). The `NY8` data set can be accessed from the **R** package `DClusterm`.  

```{r}
# Load data
data(NY8)

# View data
#head(NY8)
NY8
```

```{r}
# Check class
class(NY8)
```

```{r}
# Convert it to a df? 
# https://www.paulamoraga.com/book-geospatial/sec-spatialdataandCRS.html
NY8@data %>% head %>% pander
```

```{r}
# # Plot it 
# plot(NY8) # Just the map now.
```




## Plotting 

```{r}
# Convert to sf
library(sf)
NY8_sf <- st_as_sf(NY8)
```

```{r}
# Create the standardized mortality ratio (SMR) variable
# https://www.r-bloggers.com/2019/11/spatial-data-analysis-with-inla/
rate <- sum(NY8_sf$Cases) / sum(NY8_sf$POP8)

NY8_sf <- NY8_sf %>% mutate(
  Expected = POP8 * rate, 
  SMR = Cases / Expected
)
```

```{r}
# Plot SMR 
ggplot(NY8_sf) + geom_sf(aes(fill = SMR)) + # Look nice!
  scale_fill_gradient(high = "red") 
```




## Subsetting then plotting

```{r}
# Subset to include Syracuse city only 
syracuse <- which(NY8$AREANAME == "Syracuse city")

# Plot it
ggplot(NY8_sf[syracuse, ]) + geom_sf(aes(fill = SMR)) + 
  scale_fill_gradient(high = "red") 
```




## Poisson Models

### Fitting a Poisson regression model 

```{r}
#install.packages("INLA") # run once
#not available for this R version...
#install.packages("INLA", repos=c(getOption("repos"), INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE) # load a while
library(INLA) # Now it works.
```

Let's work on some toy examples first before coming to fix the issue. Toy examples work fine. Issues seem to related to `Cases`. Rounding `Cases` work but results differ a bit. 

```{r}
# Fit a Poisson regression model
m_fixed <- inla(round(Cases) ~ 1 + AVGIDIST,
                data = NY8_sf,
                family = "poisson",
                E = NY8_sf$Expected,
                control.predictor = list(compute = TRUE),
                control.compute = list(dic = TRUE, waic = TRUE))
```

```{r}
# summary(m_fixed) %>% pander # very bad!
```

```{r}
summary(m_fixed)
```


### Fitting a Poisson regression model with random effects 

```{r}
# Fit a Poisson regression model with random effects 
NY8_sf <- NY8_sf %>% mutate(
  ID = 1:nrow(NY8)) # Use ID as the random effect

m_random <- inla(round(Cases) ~ 1 + AVGIDIST + f(ID, model = "iid"),
                 data = NY8_sf, 
                 family = "poisson",
                 E = NY8_sf$Expected,
                 control.predictor = list(compute = TRUE),
                 control.compute = list(dic = TRUE, waic = TRUE))
```

```{r}
summary(m_random)
```


### Plotting

```{r}
# Add fitted values for both m1 & m2
NY8_sf <- NY8_sf %>% mutate(
  FIXED_EFF = m_fixed$summary.fitted[, "mean"],
  IID_EFF = m_random$summary.fitted[, "mean"]
  ) 
```

```{r}
# Plot them but for Syracuse city only
ggplot(NY8_sf[syracuse, ]) + geom_sf(aes(fill = SMR)) + 
  scale_fill_gradient(high = "red", limits = c(0, 13)) -> p_m0
p_m0
```

```{r}
ggplot(NY8_sf[syracuse, ]) + geom_sf(aes(fill = FIXED_EFF)) + #, show.legend = FALSE) +
  scale_fill_gradient(high = "red", limits = c(0, 13)) -> p_m1
p_m1 
```

```{r}
ggplot(NY8_sf[syracuse, ]) + geom_sf(aes(fill = IID_EFF)) + # , show.legend = FALSE) +
  scale_fill_gradient(high = "red", limits = c(0, 13)) -> p_m2
p_m2
```

We might want them plotted with the same scale. 

```{r}
#grid.arrange(p_m0, p_m1, p_m2, nrow = 3, ncol = 1)
```




## Spatial Models for Areal (or Lattice) Data

### Plot spatial neighbors

An adjacency (or neighbour) matrix $W$ is often used to describe spatial proximity in areal (lattice) data. Element $W_{ij}$ is non-zero, if area $i$ and $j$ are neighbors. Element $W_{ij}$ is zero, otherwise. 

```{r}
# Compute adjacency matrix
NY8.nb <- poly2nb(NY8) # construct the neighbours list / neighbour matrix
NY8.nb
```

```{r}
class(NY8.nb)
```

### Plot spatial neighbors using ggplot2 

```{r}
# Plot spatial neighbors using ggplot2 
# https://mbjoseph.github.io/posts/2018-12-27-plotting-spatial-neighbors-in-ggplot2/
NY8_sp <- as(NY8_sf, 'Spatial')  # NY8_sf is a "sf" "data.frame"
class(NY8_sp) # Now is a "SpatialPolygonsDataFrame"
```

```{r}
neighbors <- poly2nb(NY8) # construct the neighbours list
neighbors_sf <- as(nb2lines(neighbors, coords = coordinates(NY8_sp)), 'sf')
neighbors_sf <- st_set_crs(neighbors_sf, st_crs(NY8_sf))
```

```{r}
ggplot(NY8_sf) + 
  geom_sf() + # remove aes(fill = SMR)
  geom_sf(data = neighbors_sf)
```

```{r}
#plot(NY8) 
```

```{r}
# plot(NY8) 
# plot(NY8.nb, coordinates(NY8), add = TRUE, pch = ".", col = "gray")
```


```{r}
# Create sparse adjacency matrix
# Or use the function nb2INLA to generate spatial neighbours for INLA
NY8.mat <- as(nb2mat(NY8.nb, style = "B"), "Matrix") # generate a weights matrix for a neighbours list with spatial weights 
# Use this (NY8.mat) for the graph argument in the function inla
#NY8.mat
class(NY8.mat)
```

Here is [a post](https://stats.stackexchange.com/questions/445403/neighborhood-matrix-in-spatio-temporal-model) that discusses the function `poly2nb` vs. `nd2INLA`. We might also need to check [this tutorial](https://ourcodingclub.github.io/tutorials/spatial-modelling-inla/#point) to do more. 

### Plot the adjacency matrix

Is the adjacency matrix the same as spatial neighbor? 

```{r}
# Plot the adjacency matrix
image(NY8.mat)
```

```{r}
# summ <- summary(NY8.mat)
# #summ
# NY8.mat.df <- data.frame(
#   Origin = rownames(NY8.mat)[summ$i],
#   Destination = colnames(NY8.mat)[summ$j],
#   Weight = NY8.mat$x)
```




## Generalized Linear Models With Spatial Random Effects

The GLMs have the following form: 
$$
Y = X \beta + Z \alpha + \varepsilon,
$$
where $\beta$ is a vector of fixed effects with design matrix $X$, $\alpha$ is a vector of random effects with design matrix $Z$, and $\varepsilon$ is an error term, where it is assumed that $\varepsilon_i \sim N(0, \sigma^2), i = 1,...,n$. 

The vector of random effects $\alpha$ is modeled as MVN (it is assumed that) 
$$
\alpha \sim N(0, \sigma^2_\alpha \Sigma), 
$$
where $\Sigma$ is defined such that it induces higher correlation with adjacent areas. 

There are a few ways to include spatial dependence in $\Sigma$: 

1. SAR (Simultaneous autoregressive)

$$
\Sigma^{-1} = ({(I - \rho W)}^T ((I - \rho W))),
$$
where $I$ is the identity matrix, $\rho$ is a spatial autocorrelation parameter, and $W$ is the adjacency matrix. 

2. CAR (Conditional autoregressive)

$$
\Sigma^{-1} = (I - \rho W)
$$

3. ICAR (Intrinsic CAR): 

$$
\Sigma^{-1} = diag(n_i) - W,
$$
where $n_i$ is the number of neighbors of area $i$. 


4. Mixture of matrices (Leroux et al.’s model) 

$$
\Sigma^{-1} = {((1 - \lambda) I_n + \lambda M)}, \lambda \in (0, 1)
$$
where $M$ is precision of intrinsic CAR specification. 

Note. $\Sigma^{-1} = Q$ is the precision matrix. 


### Fit a SLM (spatial lag model) 

This one seems a bit complicated. Let's wait till the last or skip it altogether. 


### Fit a ICAR (Intrinsic CAR) model 

```{r}
# Setup model
# NY8.mat <- as(nb2mat(NY8.nb, style = "B"), "Matrix") # Already define earlier
```

```{r}
# Fit model
m_icar <- inla(round(Cases) ~ 1 + AVGIDIST + 
                 f(ID, model = "besag", graph = NY8.mat),
               data = NY8_sf,
               family ="poisson", 
               E = NY8_sf$Expected,
               control.predictor = list(compute = TRUE),
               control.compute = list(dic = TRUE, waic = TRUE))
summary(m_icar)
```


Later, we want to repeat these for all models. Perhaps consider a function? 

```{r}
# Get components from the results
# 2.4 Model assessment and model choice
# https://becarioprecario.bitbucket.io/inla-gitbook/ch-INLA.html#sec:modelassess
m_icar$mlik
```

```{r}
# m_icar$mlik[2,1]
m_icar$mlik[[2,1]]
```

```{r}
m_icar$dic$dic
```

```{r}
m_icar$waic$waic
```


### Fit a BYM (Besag-York-Mollié) model 

The BYM (Besag-York-Mollié) model is a convolution model of an ICAR (intrinsic CAR) effect and an iid Gaussian latent effect. 

```{r}
# Fit model
m_bym = inla(round(Cases) ~ 1 + AVGIDIST + 
               f(ID, model = "bym", graph = NY8.mat),
             data = NY8_sf, 
             family ="poisson",
             E = NY8_sf$Expected, 
             control.predictor = list(compute = TRUE),
             control.compute = list(dic = TRUE, waic = TRUE))
summary(m_bym)
```


### Fit a mixture (Leroux et al.) model

```{r}
# Setup model
ICARmatrix <- Diagonal(nrow(NY8.mat), apply(NY8.mat, 1, sum)) - NY8.mat
Cmatrix <- Diagonal(nrow(NY8), 1) - ICARmatrix
```

```{r}
# Fit model
m_ler = inla(round(Cases) ~ 1 +  AVGIDIST + 
               f(ID, model = "generic1", Cmatrix = Cmatrix), 
             data = NY8_sf,
             family ="poisson",
             E = NY8_sf$Expected, 
             control.predictor = list(compute = TRUE),
             control.compute = list(dic = TRUE, waic = TRUE))
summary(m_ler)
```

Results for all models differ a bit. 


## Get results for all models 

```{r}
# Get criteria
get_criteria <- function(model){
  mlik = model$mlik[[2,1]]
  dic = model$dic$dic
  waic = model$waic$waic
  criteria = c(mlik, dic, waic)
  return(criteria)
}
```

```{r}
# Test
x = get_criteria(m_icar)
x
```

```{r}
# # Still need SLM
# get_criteria(m_fixed)
# get_criteria(m_random)
# get_criteria(m_icar)
# get_criteria(m_bym)
# get_criteria(m_ler)
```

```{r}
# Put into a df and consider saving to the folder results
criteria_df <- tibble(
  "fixed" = get_criteria(m_fixed), 
  "iid" = get_criteria(m_random),
  "ICAR" = get_criteria(m_icar), 
  "BYM" = get_criteria(m_bym), 
  "Leroux" = get_criteria(m_ler)
)
criteria_df
```

We want DIC (deviance information criterion) and WAIC (Watanabe-Akaike information criterion (WAIC) to be low since lower DIC or WAIC value indicates better fit of the model. 

```{r}
criteria_df <- criteria_df %>% rotate_df() %>% rename(
  Marg_logLik =  V1,
  DIC = V2, 
  WAIC = V3
) # %>% pander # won't work. %>% pander later. 
```

```{r}
# Write to the results folder 
write_rds(criteria_df, here("results", "criteria_df.rds"))
```


## Plot results for all models 

Can we also create a function for plotting? 

```{r}
# Create sf for plotting
NY8_sf <- NY8_sf %>% mutate(
  FIXED_EFF = m_fixed$summary.fitted[, "mean"], 
  IID_EFF = m_random$summary.fitted[, "mean"], 
  ICAR = m_icar$summary.fitted[, "mean"], 
  BYM = m_bym$summary.fitted[, "mean"], 
  LER = m_ler$summary.fitted[, "mean"]
  ) 
```

```{r}
# Plot base case
ggplot(NY8_sf[syracuse, ]) + 
  geom_sf(aes(fill = SMR)) + 
  scale_fill_gradient(high = "red", limits = c(0, 2)) # Change limits but why? 
```

```{r}
ggplot(NY8_sf[syracuse, ]) + 
  geom_sf(aes(fill = FIXED_EFF)) + 
  scale_fill_gradient(high = "red", limits = c(0, 2))
```

```{r}
ggplot(NY8_sf[syracuse, ]) + 
  geom_sf(aes(fill = IID_EFF)) + 
  scale_fill_gradient(high = "red", limits = c(0, 2))
```

Something is off. 

```{r}
ggplot(NY8_sf[syracuse, ]) + 
  geom_sf(aes(fill = ICAR)) + 
  scale_fill_gradient(high = "red", limits = c(0, 2))
```

```{r}
ggplot(NY8_sf[syracuse, ]) + 
  geom_sf(aes(fill = BYM)) + 
  scale_fill_gradient(high = "red", limits = c(0, 2))
```

```{r}
ggplot(NY8_sf[syracuse, ]) + 
  geom_sf(aes(fill = LER)) + 
  scale_fill_gradient(high = "red", limits = c(0, 2))
```

\newpage

## Reference 

[Gómez-Rubio, V. (2019). R-bloggers. Spatial Data Analysis with INLA. https://www.r-bloggers.com/2019/11/spatial-data-analysis-with-inla/.](https://www.r-bloggers.com/2019/11/spatial-data-analysis-with-inla/)



