---
title: "00_Leukemia_in_NY"
author: "Frances Lin"
date: "3/23/2022"
output: pdf_document
---

## Load packages

```{r}
# Load packages
library(spdep)
library(DClusterm) # data
library(tidyverse)
library(pander)
library(ggplot2)
```


## The `NY8` data

The `NY8` data set contains the number of leukemia cases in an eight-country region of upstate New York from 1978-1982. 

```{r}
# Load data
data(NY8)

# View data
#head(NY8)
NY8
```

```{r}
# Check class
class(NY8)
```

```{r}
# Convert it to a df? 
# https://www.paulamoraga.com/book-geospatial/sec-spatialdataandCRS.html
head(NY8@data) %>% pander
```

```{r}
# # Plot it 
# plot(NY8) # Just the map now.
```


## Plotting 

```{r}
# Convert to sf
library(sf)
NY8_sf <- st_as_sf(NY8)
```

```{r}
# Create the standardized mortality ratio (SMR) variable
# https://www.r-bloggers.com/2019/11/spatial-data-analysis-with-inla/
rate <- sum(NY8_sf$Cases) / sum(NY8_sf$POP8)

NY8_sf <- NY8_sf %>% mutate(
  Expected = POP8 * rate, 
  SMR = Cases / Expected
)
```

```{r}
# Plot SMR 
ggplot(NY8_sf) + geom_sf(aes(fill = SMR)) + # Look nice!
  scale_fill_gradient(high="red") 
```


## Subsetting then plotting

```{r}
# Subset to include Syracuse city only 
syracuse <- which(NY8$AREANAME == "Syracuse city")

# Plot it
ggplot(NY8_sf[syracuse, ]) + geom_sf(aes(fill = SMR)) + 
  scale_fill_gradient(high="red") 
```

## Fitting a mixed-effects (Poisson regression) model 

```{r}
#install.packages("INLA") # run once
#not available for this R version...
#install.packages("INLA", repos=c(getOption("repos"), INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE) # load a while
library(INLA) # now it works?
```

Let's work on some toy examples first before coming to fix the issue. 

```{r}
# # Fit a Poisson regression model
# m1 <- inla(Cases ~ 1 + AVGIDIST,
#            data = NY8_sf,
#            family = "poisson",
#            E = NY8_sf$Expected, 
#            control.predictor = list(compute = TRUE),
#            control.compute = list(dic = TRUE, waic = TRUE))
```

```{r}
# # Fit a different model (random-effects model)
# NY8_sf <- NY8_sf %>% mutate(
#   ID = 1:nrow(NY8))
# 
# m2 <- inla(Cases ~ 1 + AVGIDIST + f(ID, model = "iid"),
#   data = as.data.frame(NY8), family = "poisson", 
#   E = NY8$Expected,
#   control.predictor = list(compute = TRUE),
#   control.compute = list(dic = TRUE, waic = TRUE))
```



\newpage


## Reference 

Gómez Rubio, Virgilio. (2019). R-bloggers. Spatial Data Analysis with INLA. [https://www.r-bloggers.com/2019/11/spatial-data-analysis-with-inla/.](https://www.r-bloggers.com/2019/11/spatial-data-analysis-with-inla/)