---
title: "01-meuse-INLA"
author: "Frances Lin"
date: "3/28/2022"
output: pdf_document
---

The objective of this project is to get familiar with and compare kriging (uk) and spde (stochastic partial differential equation) via INLA. 

## Load packages

```{r}
library(gstat) # data
library(INLA)
library(tidyverse)
library(pander)
library(sp) # will retire
library(maptools) # for unionSpatialPolygons
library(gridExtra)
library(here)
```




## Load data

```{r}
data(meuse)
```

```{r}
class(meuse)
```

```{r}
meuse %>% head(3) %>% pander
```

```{r}
meuse %>% dim
```


## Kriging

### Create a SpatialPointsDataFrame (the `sp` way) and assign CSR

```{r}
coordinates(meuse) <- ~x+y
proj4string(meuse) <- CRS("+init=epsg:28992")
```

```{r}
class(meuse)
```

### Do the same things for grid (prediction grid for meuse data)

```{r}
data(meuse.grid)
coordinates(meuse.grid) = ~x+y
proj4string(meuse.grid) <- CRS("+init=epsg:28992")
gridded(meuse.grid) = TRUE # Not sure.
```

### Compute the empirical variogram and fit a spherical variogram 

The variogram and kriging sections for this project are meant to be brief. Instead, the focus will be on expanding the INLA section. 

```{r}
vgm <- variogram(log(zinc) ~ dist, meuse)
fit.vgm <- fit.variogram(vgm, vgm("Sph"))
fit.vgm 
```

Sill (variance) = mean of psill = 0.14086. Nugget (variance when distance = 0) = min of psill = 0.07643. Range (distance when the curve first to flattens out) = mean of range = 364.3. 

```{r}
vgm_summary <- summary(fit.vgm) 
vgm_summary 
```
```{r}
# Write to the results folder 
write_rds(fit.vgm, here("results", "fit.vgm.rds"))
write_rds(vgm_summary, here("results", "vgm_summary.rds"))
```

### Fit the (universal) kriging model

```{r}
krg <- krige(log(zinc) ~ dist, meuse, meuse.grid, model = fit.vgm)
krg %>% as.data.frame %>% head %>% pander 
```

```{r}
#summary(krg)
```


### Visualize the results of UK model

```{r}
# Add estimates to meuse.grid
meuse.grid$zinc.krg <- krg$var1.pred
meuse.grid$zinc.krg.sd <- sqrt(krg$var1.var)
```

Results show that higher concentrations of (log) zinc in points closer to the Meuse river. 

```{r}
# Visualize the results of uk
library(scales) # for comma
krg %>% as.data.frame %>%
  ggplot(aes(x=x, y=y)) + geom_tile(aes(fill=var1.pred)) + coord_equal() +
  scale_fill_gradient(low = "yellow", high="red") +
  scale_x_continuous(labels=comma) + 
  scale_y_continuous(labels=comma) + # customise to add commma 
  labs(title = "uk (universal kriging) results") -> p_uk
p_uk
```


## Spatial Models using SPED (Stochastic Partial Differential Equations)

A spatial process with a Matérn covariance can be obtained as the weak solution to a stochastic partial differential equation (SPDE, Lindgren et al., 2011). 

It involves the following steps: 

1. Create a mash

2. Make the latent model 

3. Make an A matrix 

4. Organize the data

5. Estimate or predict 

### Preprocess data for INLA 

A mesh needs to be defined over the study region and it will be used to compute the approximation to the solution (i.e., the spatial process). 

### Define the boundary of the study region

```{r}
# Define the boundary
meuse.bdy <- unionSpatialPolygons(
  as(meuse.grid, "SpatialPolygons"), rep(1, length(meuse.grid))
)
```

### 1. Create a mash / define a two-dimensional mesh to define the set of basis functions

```{r}
pts <- meuse.bdy@polygons[[1]]@Polygons[[1]]@coords
# Create a mesh
mesh <- inla.mesh.2d(loc.domain = pts, 
                     max.edge = c(150, 500),
                     offset = c(100, 250) )
```

### Plot it

```{r}
# So that we can plot observations too! 
coo <- coordinates(meuse)
```

```{r}
par(mar = c(0, 0, 0, 0))
plot(mesh, asp = 1, main = "")
lines(pts, col = 3, with = 2)
points(coo, col = "red") # x & y in meuse 
```

I am still not quite sure what it is. Let's break them into steps.

```{r}
mesh$n # # of vertices 
```

```{r}
# par(mar = c(0, 0, 0, 0))
# plot(mesh, asp = 1, main = "") 
```

### Plot them side-by-side

```{r}
# par(mfrow=c(2,2)) # This turns out to be really bad. Try ggplot2 instead.
# # Plot1
# bubble(meuse, "zinc", main = "zinc concentrations (ppm)")
# # Plot2
# plot(mesh, asp = 1, main = "")
# lines(pts, col = 3, with = 2)
# points(coo, col = "red") # x & y in meuse 
```

### Plot it using ggplot2

```{r}
class(mesh)
```

I am going to skip this for now but check [this](https://sites.stat.washington.edu/peter/591/Lindgren.pdf) and [this](https://inbo.github.io/tutorials/tutorials/r_inla/spatial.pdf) later. 

```{r}
meuse %>% as.data.frame %>%
  ggplot(aes(x, y)) + geom_point(aes(size=zinc), alpha=3/4) + 
  ggtitle("Zinc Concentration (ppm)") + 
  coord_equal() -> p
```

```{r}
class(pts)
```

```{r}
plot(pts)
```

```{r}
pts %>% as.data.frame %>%
  ggplot(aes(x, y)) + geom_path() #geo_line doesn't work.
```

```{r}
# Convert to df for plotting
pts_df <- pts %>% as.data.frame
meuse_df <- meuse %>% as.data.frame
```

```{r}
library(inlabru)
# Use neither as the default 
ggplot(NULL, aes(x, y)) + 
  geom_point(data = meuse_df, aes(size=zinc, alpha=3/4)) + 
  geom_path(data = pts_df) + 
  gg(mesh) +
  ggtitle("with Mesh") +
  coord_equal() -> p_mesh
p_mesh
```

```{r}
# library(inlabru)
# meuse %>% as.data.frame %>% # Boundary line left to add. 
#   ggplot(aes(x, y)) + geom_point(aes(size=zinc), alpha=3/4) + 
#   gg(mesh) +
#   ggtitle("with Mesh") +
#   coord_equal() -> p_mesh
# p_mesh
```

```{r}
# Plot them 
grid.arrange(p, p_mesh, nrow = 2)
```

### 2. Make the latent model / create a SPDE (stochastic partial differential equation) model

A spatial process with a Matérn covariance can be obtained as the weak solution to a stochastic partial differential equation (SPDE) (Lindgren et al., 2011). 

```{r}
# Build a SPDE model 
meuse.spde <- inla.spde2.matern(mesh = mesh, alpha = 2) # alpha = smoothness parameter 

# Generate the index set for the SPDE model 
s.index <- inla.spde.make.index(name = "spatial.field",
                                n.spde = meuse.spde$n.spde)
```


### 3. Make an A matrix 

```{r}
# Construct a projection matrix A to project the GRF from the observations to the triangulation vertices 
A.meuse <- inla.spde.make.A(mesh = mesh, loc = coordinates(meuse)) # Use meuse for estimation 

# Construct another projection matrix A for prediction
A.pred <- inla.spde.make.A(mesh = mesh, loc = coordinates(meuse.grid)) # Use meuse.grid for prediction 
```


### 4. Organize the data used for estimation (model fitting) or prediction 

```{r}
# Organize the data, projection matrices and effects for estimation 
meuse.stack <- inla.stack(data = list(zinc = meuse$zinc),
                          A = list(A.meuse, 1),
                          effects = list(c(s.index, list(Intercept = 1)),
                                         list(dist = meuse$dist)),
                          tag = "meuse.data") # Change to meuse.est? 

# Organize the data for prediction 
meuse.stack.pred <- inla.stack(data = list(zinc = NA),
                               A = list(A.pred, 1),
                               effects = list(c(s.index, list (Intercept = 1)),
                                              list(dist = meuse.grid$dist)),
                               tag = "meuse.pred")
```


### Join stacks of data into a single object 

```{r}
join.stack <- inla.stack(meuse.stack, meuse.stack.pred)
```


### Fit the spatial model

Other models that can go into the `model =` argument in `f()` include, for example, `spde` (Matèrn correlation (continuous)), `matern2d` (Matèrn correlation (discrete)), `besag` (Intrinsic CAR), `besagproper` (Proper CAR), and `bym` (Convolution). We will explore this more in the future project. 

```{r}
# Specify the formula
form <- log(zinc) ~ -1 + Intercept + dist + f(spatial.field, model = spde) # - 1 removes intercept 
```

```{r}
# Fit the model
start.time <- Sys.time()
m1 <- inla(form, 
           data = inla.stack.data(join.stack, spde = meuse.spde),
           family = "gaussian",
           control.predictor = list(A = inla.stack.A(join.stack), compute = TRUE),
           control.compute = list(cpo = TRUE, dic = TRUE))
end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken # Time difference of 9.82935 secs
```

According to [this site](https://www.paulamoraga.com/book-geospatial/sec-inla.html), the main arguments of the `inla()` function are 

- `formula`: 

- `data`: 

- `family`: 

- `control.predictor`: 

- `control.compute`: .


### Print summary results 

```{r}
summary(m1)
```

### Extract summary of the fixed effects

```{r}
m1$summary.fixed
```


### Extract summary of the random effects

```{r}
#m1$summary.random
```

### Extract summary of the hyperparameters

```{r}
m1$summary.hyperpar
```

### Extract results for comparision 

```{r}
# Compute statistics in terms or range and variance
spde.est <- inla.spde2.result(inla = m1, 
                              name = "spatial.field",
                              spde = meuse.spde, 
                              do.transf = TRUE)
```

Values differ a bit. 

```{r}
# Variance to compare to sill 
variance <- inla.zmarginal(spde.est$marginals.variance.nominal[[1]])
variance
```
```{r}
# Range
range <- inla.zmarginal(spde.est$marginals.range.nominal[[1]])
range
```

```{r}
# Write to the results folder 
write_rds(variance, here("results", "variance.rds"))
write_rds(range, here("results", "range.rds"))
```

The `inlabru` package (Bachl et al., 2019) can simplify the way in which the model is defined and fit. 

### Plot it 

```{r}
class(krg)
```

```{r}
class(m1)
```

```{r}
#krg
```

```{r}
#m1
```

This might not work but I am gonna try. 

```{r}
spde <- m1
```

```{r}
#spde %>% as.data.frame # Does not work.
```

```{r}
# Load spatial domain to interpolate over
data("meuse.grid") # What is this???
```

This section needs to be rewritten to be more organized. 

```{r}
# For krg 
meuse.grid$zinc.krg <- krg$var1.pred
```

```{r}
# For spde
# Obtain the indices of thr rows corresponding to the predictions 
index.pred <- inla.stack.index(join.stack, tag = "meuse.pred")$data
# Create a variable zinc.spde with the posterior mean 
meuse.grid$zinc.spde <- spde$summary.fitted.values[index.pred, "mean"]
```

```{r}
meuse.grid$zinc.spde.ll <- spde$summary.fitted.values[index.pred, "0.025quant"] # lower limit of 95% credible intervals 
meuse.grid$zinc.spde.ul <- spde$summary.fitted.values[index.pred, "0.975quant"]
```

```{r}
meuse.grid %>% head(3) %>% pander
```

```{r}
krg %>% as.data.frame %>%
  ggplot(aes(x=x, y=y)) + geom_tile(aes(fill=var1.pred)) + coord_equal() +
  scale_fill_gradient(low = "yellow", high="red") +
  scale_x_continuous(labels=comma) + 
  scale_y_continuous(labels=comma) + # customize to add commma 
  labs(title = "uk (universal kriging) results") -> p_uk
#p_uk
```

```{r}
# Plot for krg
meuse.grid %>% as.data.frame %>% 
  ggplot(aes(x=x, y=y)) + geom_tile(aes(fill=zinc.krg)) + coord_equal() +
  scale_fill_gradient(low = "yellow", high="red") +
  scale_x_continuous(labels=comma) + 
  scale_y_continuous(labels=comma) +
  labs(title = "uk results") -> p_krg
#p_krg # This should match the plot from above.
```

```{r}
# Plot for spde
meuse.grid %>% as.data.frame %>% 
  ggplot(aes(x=x, y=y)) + geom_tile(aes(fill=zinc.spde)) + coord_equal() +
  scale_fill_gradient(low = "yellow", high="red") +
  scale_x_continuous(labels=comma) + 
  scale_y_continuous(labels=comma) +
  labs(title = "spde results") -> p_spde
#p_spde
```

### Results of spde with credible intervals 

```{r}
# For ll 
meuse.grid %>% as.data.frame %>% 
  ggplot(aes(x=x, y=y)) + geom_tile(aes(fill=zinc.spde.ll)) + coord_equal() +
  scale_fill_gradient(low = "yellow", high="red") +
  scale_x_continuous(labels=comma) + 
  scale_y_continuous(labels=comma) +
  labs(title = "spde results ll") -> p_spde_ll

# For ul 
meuse.grid %>% as.data.frame %>% 
  ggplot(aes(x=x, y=y)) + geom_tile(aes(fill=zinc.spde.ul)) + coord_equal() +
  scale_fill_gradient(low = "yellow", high="red") +
  scale_x_continuous(labels=comma) + 
  scale_y_continuous(labels=comma) +
  labs(title = "spde results ul") -> p_spde_ul
```

```{r}
grid.arrange(p_spde_ll, p_spde, p_spde_ul, ncol = 2)
```


We will come back to explore more and add additional details. 


## Compare results of uk vs spde

### Plot results of uk vs spde

```{r}
# Plot them side-by-side
grid.arrange(p_krg, p_spde, ncol = 2)
```

### Plot difference plot of uk vs spde

Areas where it's colored in either red or blue are where `krg` results & `spde` results differ the most since we set `zinc.diff` = `krg` results - `spde` results. 

```{r}
p_diff <- meuse.grid %>% as.data.frame %>% 
  mutate(
    zinc.diff = zinc.krg - zinc.spde
  ) %>% 
  ggplot(aes(x=x, y=y)) + geom_tile(aes(fill=zinc.diff)) + coord_equal() +
  scale_fill_gradient(low = "blue", high="red") +
  scale_x_continuous(labels=comma) + 
  scale_y_continuous(labels=comma) +
  labs(title = "difference of uk vs spde results") 
p_diff
```

```{r}
# Write to the results folder 
write_rds(p_krg, here("results", "p_krg.rds"))
write_rds(p_spde, here("results", "p_spde.rds"))
write_rds(p_diff, here("results", "p_diff.rds"))
```

\newpage

## Reference 

[Gómez-Rubio, V. (2020). 7.3 Geostatistics. *Bayesian inference with INLA.* CRC Press.](https://becarioprecario.bitbucket.io/inla-gitbook/ch-spatial.html#sec:geostats)

[Moraga, P. (2019). 8 Geostatistical data. *Geospatial health data: Modeling and visualization with R-INLA and shiny.* CRC Press.](https://www.paulamoraga.com/book-geospatial/sec-geostatisticaldatatheory.html#spatial-modeling-of-rainfall-in-paraná-brazil)



